---
title: "Make Countries Move"
author: "Zaid Khalaf"
date: "14/03/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

#  Explore global development with R
In this exercise, you will load a filtered `gapminder` dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks within this script.** 

- **Submit to Github this script with solutions and its knitted version.**  
- **Submit to Brightspace a document with the (1) URL of this script and its knitted version in Github (2) answer of 250 words and a (3) visualisation of the Final Question**

## Get the necessary packages
First, start with installing and activating the relevant packages `tidyverse`, `gganimate`, and `gapminder` if you do not have them already. 
*Pay attention to what warning messages you get when installing `gganimate`, as your computer might need other packages than `gifski` and `av`*

```{r libraries, echo = FALSE}
# Set CRAN mirror to a default repository (for example, RStudio's server)
options(repos = c(CRAN = "https://cran.rstudio.com"))

install.packages("gganimate")
install.packages("gifski")
install.packages("av")
install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)

```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
head(gapminder)


highest_gdp_country <- subset(gapminder, year == 1952)[which.max(subset(gapminder, year == 1952)$gdpPercap), ]
highest_gdp_country

gdp_country_2007 <- subset(gapminder, year==2007)

gdp_2007_sorted <- gdp_country_2007[order(-gdp_country_2007$gdpPercap), ]

Five_richest_countries <- head(gdp_2007_sorted, 5)

Five_richest_countries
```


The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point(aes(color=continent)) +
  scale_x_log10() +
  ggtitle("Figure 01")+
  xlab("GDP per Citizen")+
  ylab("Life Expectancy")
  options(scipen = 999)

```

...

We see an interesting spread with an outlier to the right. Explore who it is so you can answer question 2 below!


Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point(aes(color=continent)) +
  scale_x_log10() +
  ggtitle("Figure 02")+ 
  xlab("GDP per Citizen") +
  ylab("Life Expectancy") 
options(scipen = 999)

```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

### Questions for the static figures:

1. *Answer: why does it make sense to have a log10 scale (`scale_x_log10()`) on the x axis?* **Answer:** A log scale on the x-axis compresses the wide range of GDP values, preventing high-GDP outliers from overshadowing lower-GDP countries and revealing meaningful patterns across all income levels.

2. *Answer: In Figure 1: Who is the outlier (the richest country in 1952) far right on the x axis?* Kuwait

3. *Fix Figures 1 and 2: Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation". You want to eliminate it.)* aes(color = continent).

4. *Answer: What are the five richest countries in the world in 2007?*
Norway	
Kuwait	
Singapore	
United States	Americas	
Ireland	Europe



##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point(aes(color=continent)) +
  scale_x_log10()  # convert x to log scale
anim +
  xlab("GDP per Citizen") +
  ylab("Life Expectancy") 
options(scipen = 999)

```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1) +
  xlab("GDP per Citizen") +
  ylab("Life Expectancy") 
```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smooths the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +  
  scale_size(range = c(2, 12)) +
  labs(title = "Year: {frame_time}",  
       x = "GDP per Capita (log scale)",
       y = "Life Expectancy (years)",
       size = "Population",
       color = "Continent") +
  theme_minimal() +
  theme(
    legend.position = "right",
    text = element_text(size = 14)  
  ) +
  transition_time(year) +  
  ease_aes('linear')  


ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +  
  scale_size(range = c(2, 12)) +
  labs(title = "GDP per Capita vs Life Expectancy in {closest_state}",
       x = "GDP per Capita (log scale)",
       y = "Life Expectancy (years)",
       size = "Population",
       color = "Continent") +
  theme_minimal() +
  transition_states(year, transition_length = 1, state_length = 1)



# Animate using transition_time
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +  
  scale_y_continuous(labels = scales::comma) +  
  scale_size(range = c(2, 12)) +  
  labs(title = "Year: {frame_time}",  
       x = "GDP per Capita (log scale)",  
       y = "Life Expectancy (years)",  
       size = "Population",  
       color = "Continent") + 
  theme_minimal() +
  theme(
    legend.position = "right",  
    text = element_text(size = 14)  
  ) +
  transition_time(year) 
  ease_aes('linear')



  birth_year <- 1982

gapminder %>%
  filter(year %in% c(birth_year, 2007)) %>%
  group_by(year) %>%
  summarise(
    avg_lifeExp = mean(lifeExp, na.rm = TRUE),
    avg_gdpPercap = mean(gdpPercap, na.rm = TRUE),
    avg_pop = mean(pop, na.rm = TRUE)
  )


ggplot(gapminder %>% filter(year %in% c(birth_year, 2007)), 
       aes(x = factor(year), y = lifeExp, fill = continent)) +
  geom_boxplot() +
  labs(title = "Life Expectancy in My Birth Year vs 2007",
       x = "Year",
       y = "Life Expectancy (years)",
       fill = "Continent") +
  theme_minimal()

```


The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 

### Tasks for the animations:

5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)* I added labs(title = "Year: {frame_time}")


6. *Can you made the axes' labels and units more readable? Consider expanding the abbreviated labels as well as the scientific notation in the legend and x axis to whole numbers. Also, differentiate the countries from different continents by color*



### Final Question
7. **Is the world a better place today than it was in the year you were born?** Based on the data from 1982 and 2007, we can see that the world has become a better place in several ways, but not for everyone equally. First, looking at life expectancy, people in 2007 lived longer than in 1982. The average life expectancy in 1982 was around 60 years, while in 2007, it was about 67 years. This means that healthcare, better living conditions, and advancements in medicine have helped people live longer lives. Next, when we look at GDP per capita, which is how much money a person makes on average, it shows that the world has gotten richer. In 1982, the average GDP per capita was lower than in 2007. This shows that many countries became wealthier over time, and more people were able to afford better things like food, housing, and education.However, not all countries improved at the same rate. While some countries like the United States, Norway, and Japan became wealthier and had better healthcare, some countries in Africa, for example, still faced challenges like poverty and disease.
In conclusion, the world has become more prosperous and healthier since 1982, but the benefits haven't been shared equally. Some countries have improved a lot, but others still face challenges that need to be addressed for everyone to have a better life.


